Jane, John … Leslie? A Historical Method for Algorithmic Gender Prediction

### 2014-10-07 article.Rmd
Formatting figures for submission

---
title: "Jane, John … Leslie? A Historical Method for Algorithmic Gender Prediction"
author: "Cameron Blevins and Lincoln Mullen"
output: html_document
---

@@

article_palette <- "Dark2"
f_color <- "#8E001C"
u_color <- "#D3CEAA"
m_color <- "#4C4C4C"
tri_color <- c(f_color, u_color, m_color) 
```

@@

```{r plot-leslie}
# We cannot simply use the Social Security figures (even though they probably
# don't present a particular problem for the name Leslie) because later we will
# argue that the gender ratios for the SSA figures before 1930 are skewed. So we
# are going to splice the IPUMS data from before 1930 and the SSA data from
# after 1930.

@@

leslie_df %>%
  ggplot(aes(x = year, y = prop_female)) +
  geom_smooth() + 
  geom_point() +
  ggtitle("Gender of the name Leslie, 1820-2012") +
  xlab(NULL) + ylab("Proportion of female uses") + ylim(0, 1) +
  geom_hline(yintercept = 0.5) + 
  annotate("text", x = 1900, y = 0.45, label = "Male") +
  annotate("text", x = 1900, y = 0.55, label = "Female") +
  article_theme
```

@@

```{r plot-names-that-change}
plot_names <-  c("madison",
                 #"keanu", #keanu doesn't seem to actually flip, or at least doesn't show up on the chart?
                 "addison",
                 "jan",
                 "sydney",
                 "morgan",
                 "leslie",
                 "kendall")

ssa_national %>%
  filter(name %in% plot_names) %>%
  mutate(prop_female = female / (female + male)) %>%
ggplot(aes(x = year, y = prop_female, color = name)) +
  geom_smooth(se = FALSE) +
  ggtitle("Changes in gender of selected names, 1930-2012") +
  xlim(1930, 2012) + 
  xlab(NULL) + ylab("Proportion of female uses") + ylim(0, 1) +
  geom_hline(yintercept = 0.5) +
  annotate("text", x = 1965, y = 0.45, label = "Male") +
  annotate("text", x = 1965, y = 0.55, label = "Female") +
  article_theme +
  scale_color_brewer(palette = article_palette, name = "")
```

@@

```{r}
genderdata::ssa_national %>%
  group_by(year) %>%
  summarize(ratio = sum(female) /sum(female + male)) %>%
  ggplot(aes(x = year, y = ratio)) +
  geom_line() +
  ggtitle("Percentage of SSA dataset that is female") +
  xlab(NULL) + ylab(NULL) +
  article_theme
```

@@

```{r}
diss_predicted %>%
  mutate(gender_author = factor(gender_author, 
                                levels = c("female", "unknown", "male"))) %>%
  ggplot(aes(x = year, fill = gender_author)) + 
  geom_bar(binwidth = 1) +
  ggtitle("Dissertations about history by gender, 1950-2012") +
  xlab(NULL) + ylab(NULL) +
  short_scale + theme(legend.position = "bottom") +
  scale_fill_manual(values=tri_color, name = "",
                    labels = c("Female", "Unknown", "Male")) +
  guides(fill = guide_legend(reverse = TRUE)) +
  #scale_fill_brewer(palette = article_palette, name = "",
  #                    labels = c("Female", "Unknown", "Male")) +
  article_theme
```

@@

```{r}
diss_proportions <- diss_predicted %>%
  select(year, gender_author) %>%
  filter(year >= 1950) %>%
  group_by(year, gender_author) %>% 
  summarize(n = n()) %>% 
  spread(gender_author, n, fill = 0) %>% 
  mutate(female_proportion = female / (female + male + unknown),
         male_proportion = male / (female + male + unknown),
         unknown_proportion = unknown / (female + male + unknown)) %>% 
  select(female_proportion, male_proportion, unknown_proportion, year) %>%
  gather(gender, proportion, -year)

@@

diss_proportions  %>%
  ggplot(aes(x = year, y = proportion, fill = gender, order = gender)) +
  geom_bar(binwidth = 1, stat = "identity") +
  ggtitle("Proportion of dissertations about history by gender 1950-2012") +
  xlab(NULL) + ylab(NULL) +
  short_scale + theme(legend.position = "bottom") +
  geom_hline(yintercept = 0.5) +
  scale_fill_manual(values=tri_color, name = "",
                    labels = c("Female", "Unknown", "Male")) +
  guides(fill = guide_legend(reverse = TRUE)) +
  #scale_fill_brewer(palette = article_palette,
  #                  labels = c("Female", "Unknown", "Male")) +
  article_theme
```

@@

```{r}
ahr_proportions <- ahr_predicted %>%
  select(year, gender_author) %>%  
  group_by(year, gender_author) %>% 
  summarize(n = n()) %>% 
  spread(gender_author, n, fill = 0) %>%
  mutate(female_proportion = female / (female + male + unknown),
         male_proportion = male / (female + male + unknown),
         unknown_proportion = unknown / (female + male + unknown)) %>%  
  select(female_proportion, male_proportion, unknown_proportion, year) %>%
  gather(gender, proportion, -year)

@@

ahr_proportions %>%
  ggplot(aes(x = year, y = proportion, fill = gender, order = gender)) + 
  geom_bar(binwidth = 1, stat = "identity") +
  ggtitle("Authors of books reviewed in the AHR by gender") +
  xlab(NULL) + ylab(NULL) +
  theme(legend.position = "bottom") +
  geom_hline(yintercept = 0.5) +
  scale_fill_manual(values=tri_color, name = "",
                    labels = c("Female", "Unknown", "Male")) +
  guides(fill = guide_legend(reverse = TRUE)) +
  #scale_fill_brewer(palette = article_palette,
  #                    labels = c("Female", "Unknown", "Male")) +
  article_theme
```

@@

```{r}
ahr_proportions_no_unknown <- ahr_predicted %>%
  select(year, gender_author) %>%  
  group_by(year, gender_author) %>% 
  summarize(n = n()) %>% 
  spread(gender_author, n, fill = 0) %>%
  mutate(female_proportion = female / (female + male),
         male_proportion = male / (female + male)) %>%
  select(female_proportion, male_proportion, year) %>%
  gather(gender, proportion, -year)

@@

combined_proportions %>%
  ggplot(aes(x = year, y = proportion, color = gender, size = type)) + 
  geom_line() +
  scale_x_continuous(breaks = seq(1950, 2010, 10),
                     labels = seq(1950, 2010, 10),
                     limits = c(1950, 2025)) +
  scale_size_discrete(name = "Type",
                      labels = c("Book", "Dissertation"), range = c(2,1)) +
  scale_color_manual(values=c(f_color, m_color), name = "Gender",
                    labels = c("Female", "Male")) +
  guides(color=FALSE, size=FALSE) +
  #guides(color = guide_legend(reverse = TRUE)) +
  annotate("text", x = 2017, y = 0.77, label = "Male Authors\n (Books)", size=5, fontface="bold", color = m_color) +
  annotate("text", x = 2020, y = 0.57, label = "Male Authors\n (Dissertations)", size=4, color = m_color) +
  annotate("text", x = 2017, y = 0.23, label = "Female Authors\n (Books)", size=5, fontface="bold", color = f_color) +
  annotate("text", x = 2020, y = 0.43, label = "Female Authors\n (Dissertations)", size=4, color = f_color) +
  #scale_color_brewer(palette = article_palette, name = "Gender",
  #                     labels = c("Female", "Male")) +
  xlab(NULL) + ylab("Proportion of authors by gender") +
  ggtitle("Gender Gap of Dissertations and of Books Reviewed in the AHR") +
  article_theme +
  geom_hline(yintercept = 0.5)
```

@@

```{r}
ahr_predicted %>%
  filter(gender_author == "male") %>%
  select(year, gender_reviewer) %>%  
  group_by(year, gender_reviewer) %>% 
  summarize(n = n()) %>% 
  spread(gender_reviewer, n, fill = 0) %>%
  mutate(female_proportion = female / (female + male + unknown),
         male_proportion = male / (female + male + unknown),
         unknown_proportion = unknown / (female + male + unknown)) %>%  
  select(female_proportion, male_proportion, unknown_proportion, year) %>%
  gather(gender, proportion, -year) %>%
  mutate(gender = factor(gender, levels = c("female_proportion",
                                            "unknown_proportion",
                                            "male_proportion"))) %>%
  ggplot(aes(x = year, y = proportion, fill = gender, order = gender)) + 
  geom_bar(binwidth = 1, stat = "identity") +
  ggtitle("Gender of reviewers of MALE book authors in AHR") +
  xlab(NULL) + ylab(NULL) +
  theme(legend.position = "bottom") +
  geom_hline(yintercept = 0.5) +
  scale_fill_manual(values=tri_color, name = "",
                    labels = c("Female", "Unknown", "Male")) +
  guides(fill = guide_legend(reverse = TRUE)) +
  #scale_fill_brewer(palette = article_palette, 
  #                  labels = c("Female", "Unknown", "Male")) +
  article_theme +
  short_scale
```

```{r}
ahr_predicted %>%
  filter(gender_author == "female") %>%
  select(year, gender_reviewer) %>%  
  group_by(year, gender_reviewer) %>% 
  summarize(n = n()) %>% 
  spread(gender_reviewer, n, fill = 0) %>%
  mutate(female_proportion = female / (female + male + unknown),
         male_proportion = male / (female + male + unknown),
         unknown_proportion = unknown / (female + male + unknown)) %>%  
  select(female_proportion, male_proportion, unknown_proportion, year) %>%
  gather(gender, proportion, -year) %>%
  mutate(gender = factor(gender, levels = c("female_proportion",
                                            "unknown_proportion",
                                            "male_proportion"))) %>%
  ggplot(aes(x = year, y = proportion, fill = gender, order = gender)) + 
  geom_bar(binwidth = 1, stat = "identity") +
  ggtitle("Gender of reviewers of FEMALE book authors in AHR") +
  xlab(NULL) + ylab(NULL) +
  theme(legend.position = "bottom") +
  geom_hline(yintercept = 0.5) +
  scale_fill_manual(values=tri_color, name = "",
                    labels = c("Female", "Unknown", "Male")) +
  guides(fill = guide_legend(reverse = TRUE)) +
  #scale_fill_brewer(palette = article_palette, 
  #                  labels = c("Female", "Unknown", "Male")) +
  article_theme +
  short_scale
```

>>>>

---
title: "Jane, John … Leslie? A Historical Method for Algorithmic Gender Prediction"
author: "Cameron Blevins and Lincoln Mullen"
output:
  html_document: default
---

@@

article_palette <- "Dark2"
f_color <- "#8E001C"
u_color <- "#D3CEAA"
m_color <- "#4C4C4C"
tri_color <- c(f_color, u_color, m_color)

@@

```{r figure01}
# We cannot simply use the Social Security figures (even though they probably
# don't present a particular problem for the name Leslie) because later we will
# argue that the gender ratios for the SSA figures before 1930 are skewed. So we
# are going to splice the IPUMS data from before 1930 and the SSA data from
# after 1930.

@@

leslie_df %>%
  ggplot(aes(x = year, y = prop_female)) +
  geom_smooth() + 
  geom_point() +
  ggtitle("Gender of the name Leslie, 1820-2012") +
  xlab(NULL) + ylab("Proportion of female uses") + ylim(0, 1) +
  geom_hline(yintercept = 0.5) + 
  annotate("text", x = 1900, y = 0.45, label = "Male") +
  annotate("text", x = 1900, y = 0.55, label = "Female") +
  article_theme
```
**Figure 1: Gender of the name Leslie, 1820-2012**

@@

```{r figure02}
plot_names <-  c("madison",
                 #"keanu", #keanu doesn't seem to actually flip, or at least doesn't show up on the chart?
                 "addison",
                 "jan",
                 "sydney",
                 "morgan",
                 "leslie",
                 "kendall")

ssa_national %>%
  filter(name %in% plot_names) %>%
  mutate(prop_female = female / (female + male)) %>%
ggplot(aes(x = year, y = prop_female, color = name)) +
  geom_smooth(se = FALSE) +
  ggtitle("Changes in gender of selected names, 1930-2012") +
  xlim(1930, 2012) + 
  xlab(NULL) + ylab("Proportion of female uses") + ylim(0, 1) +
  geom_hline(yintercept = 0.5) +
  annotate("text", x = 1965, y = 0.45, label = "Male") +
  annotate("text", x = 1965, y = 0.55, label = "Female") +
  article_theme +
  scale_color_brewer(palette = article_palette, name = "")
```
**Figure 2: Proportion of female uses**

@@

```{r figure03}
genderdata::ssa_national %>%
  group_by(year) %>%
  summarize(ratio = sum(female) /sum(female + male)) %>%
  ggplot(aes(x = year, y = ratio)) +
  geom_line() +
  ggtitle("Percentage of SSA dataset that is female") +
  xlab(NULL) + ylab(NULL) +
  article_theme
```
**Figure 3: Percentage of SSA dataset that is female**

@@

```{r figure04}
diss_predicted %>%
  mutate(gender_author = factor(gender_author, 
                                levels = c("female", "unknown", "male"))) %>%
  ggplot(aes(x = year, fill = gender_author)) + 
  geom_bar(binwidth = 1) +
  ggtitle("Dissertations about history by gender, 1950-2012") +
  xlab(NULL) + ylab(NULL) +
  short_scale + theme(legend.position = "bottom") +
  scale_fill_manual(values=tri_color, name = "",
                    labels = c("Female", "Unknown", "Male")) +
  guides(fill = guide_legend(reverse = TRUE)) +
  #scale_fill_brewer(palette = article_palette, name = "",
  #                    labels = c("Female", "Unknown", "Male")) +
  article_theme
```
**Figure 4: Dissertations about history by gender, 1950-2012**

@@

```{r figure05}
diss_proportions <- diss_predicted %>%
  select(year, gender_author) %>%
  filter(year >= 1950) %>%
  group_by(year, gender_author) %>% 
  summarize(n = n()) %>% 
  spread(gender_author, n, fill = 0) %>% 
  mutate(female_proportion = female / (female + male + unknown),
         male_proportion = male / (female + male + unknown),
         unknown_proportion = unknown / (female + male + unknown)) %>% 
  select(female_proportion, male_proportion, unknown_proportion, year) %>%
  gather(gender, proportion, -year)

@@

diss_proportions  %>%
  ggplot(aes(x = year, y = proportion, fill = gender, order = gender)) +
  geom_bar(binwidth = 1, stat = "identity") +
  ggtitle("Proportion of dissertations about history by gender 1950-2012") +
  xlab(NULL) + ylab(NULL) +
  short_scale + theme(legend.position = "bottom") +
  geom_hline(yintercept = 0.5) +
  scale_fill_manual(values=tri_color, name = "",
                    labels = c("Female", "Unknown", "Male")) +
  guides(fill = guide_legend(reverse = TRUE)) +
  #scale_fill_brewer(palette = article_palette,
  #                  labels = c("Female", "Unknown", "Male")) +
  article_theme
```
**Figure 5: Proportion of dissertations about history by gender 1950-2012**

@@

```{r figure06}
ahr_proportions <- ahr_predicted %>%
  select(year, gender_author) %>%  
  group_by(year, gender_author) %>% 
  summarize(n = n()) %>% 
  spread(gender_author, n, fill = 0) %>%
  mutate(female_proportion = female / (female + male + unknown),
         male_proportion = male / (female + male + unknown),
         unknown_proportion = unknown / (female + male + unknown)) %>%  
  select(female_proportion, male_proportion, unknown_proportion, year) %>%
  gather(gender, proportion, -year)

@@

ahr_proportions %>%
  ggplot(aes(x = year, y = proportion, fill = gender, order = gender)) + 
  geom_bar(binwidth = 1, stat = "identity") +
  ggtitle("Authors of books reviewed in the AHR by gender") +
  xlab(NULL) + ylab(NULL) +
  theme(legend.position = "bottom") +
  geom_hline(yintercept = 0.5) +
  scale_fill_manual(values=tri_color, name = "",
                    labels = c("Female", "Unknown", "Male")) +
  guides(fill = guide_legend(reverse = TRUE)) +
  #scale_fill_brewer(palette = article_palette,
  #                    labels = c("Female", "Unknown", "Male")) +
  article_theme
```
**Figure 6: Authors of books reviewed in the AHR by gender**

@@

```{r figure07}
ahr_proportions_no_unknown <- ahr_predicted %>%
  select(year, gender_author) %>%  
  group_by(year, gender_author) %>% 
  summarize(n = n()) %>% 
  spread(gender_author, n, fill = 0) %>%
  mutate(female_proportion = female / (female + male),
         male_proportion = male / (female + male)) %>%
  select(female_proportion, male_proportion, year) %>%
  gather(gender, proportion, -year)

@@

combined_proportions %>%
  ggplot(aes(x = year, y = proportion, color = gender, size = type)) + 
  geom_line() +
  scale_x_continuous(breaks = seq(1950, 2010, 10),
                     labels = seq(1950, 2010, 10),
                     limits = c(1950, 2025)) +
  scale_size_discrete(name = "Type",
                      labels = c("Book", "Dissertation"), range = c(2,1)) +
  scale_color_manual(values=c(f_color, m_color), name = "Gender",
                    labels = c("Female", "Male")) +
  guides(color=FALSE, size=FALSE) +
  #guides(color = guide_legend(reverse = TRUE)) +
  annotate("text", x = 2017, y = 0.77, label = "Male Authors\n (Books)", size=5, fontface="bold", color = m_color) +
  annotate("text", x = 2020, y = 0.57, label = "Male Authors\n (Dissertations)", size=4, color = m_color) +
  annotate("text", x = 2017, y = 0.23, label = "Female Authors\n (Books)", size=5, fontface="bold", color = f_color) +
  annotate("text", x = 2020, y = 0.43, label = "Female Authors\n (Dissertations)", size=4, color = f_color) +
  #scale_color_brewer(palette = article_palette, name = "Gender",
  #                     labels = c("Female", "Male")) +
  xlab(NULL) + ylab("Proportion of authors by gender") +
  ggtitle("Gender Gap of Dissertations and of Books Reviewed in the AHR") +
  article_theme +
  geom_hline(yintercept = 0.5)
```
**Figure 7: Gender Gap of Dissertations and of Books Reviewed in the *AHR***

@@

```{r figure08}
ahr_predicted %>%
  filter(gender_author == "male") %>%
  select(year, gender_reviewer) %>%  
  group_by(year, gender_reviewer) %>% 
  summarize(n = n()) %>% 
  spread(gender_reviewer, n, fill = 0) %>%
  mutate(female_proportion = female / (female + male + unknown),
         male_proportion = male / (female + male + unknown),
         unknown_proportion = unknown / (female + male + unknown)) %>%  
  select(female_proportion, male_proportion, unknown_proportion, year) %>%
  gather(gender, proportion, -year) %>%
  mutate(gender = factor(gender, levels = c("female_proportion",
                                            "unknown_proportion",
                                            "male_proportion"))) %>%
  ggplot(aes(x = year, y = proportion, fill = gender, order = gender)) + 
  geom_bar(binwidth = 1, stat = "identity") +
  ggtitle("Gender of reviewers of MALE book authors in AHR") +
  xlab(NULL) + ylab(NULL) +
  theme(legend.position = "bottom") +
  geom_hline(yintercept = 0.5) +
  scale_fill_manual(values=tri_color, name = "",
                    labels = c("Female", "Unknown", "Male")) +
  guides(fill = guide_legend(reverse = TRUE)) +
  #scale_fill_brewer(palette = article_palette, 
  #                  labels = c("Female", "Unknown", "Male")) +
  article_theme +
  short_scale
```
**Figure 8: Gender of reviewers of MALE book authors in *AHR***

```{r figure09}
ahr_predicted %>%
  filter(gender_author == "female") %>%
  select(year, gender_reviewer) %>%  
  group_by(year, gender_reviewer) %>% 
  summarize(n = n()) %>% 
  spread(gender_reviewer, n, fill = 0) %>%
  mutate(female_proportion = female / (female + male + unknown),
         male_proportion = male / (female + male + unknown),
         unknown_proportion = unknown / (female + male + unknown)) %>%  
  select(female_proportion, male_proportion, unknown_proportion, year) %>%
  gather(gender, proportion, -year) %>%
  mutate(gender = factor(gender, levels = c("female_proportion",
                                            "unknown_proportion",
                                            "male_proportion"))) %>%
  ggplot(aes(x = year, y = proportion, fill = gender, order = gender)) + 
  geom_bar(binwidth = 1, stat = "identity") +
  ggtitle("Gender of reviewers of FEMALE book authors in AHR") +
  xlab(NULL) + ylab(NULL) +
  theme(legend.position = "bottom") +
  geom_hline(yintercept = 0.5) +
  scale_fill_manual(values=tri_color, name = "",
                    labels = c("Female", "Unknown", "Male")) +
  guides(fill = guide_legend(reverse = TRUE)) +
  #scale_fill_brewer(palette = article_palette, 
  #                  labels = c("Female", "Unknown", "Male")) +
  article_theme +
  short_scale
```
**Figure 9: Gender of reviewers of FEMALE book authors in *AHR***
